---
title: "Epidemic simulation and inference from the simulated data"
author: "Lucy M Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, warning=FALSE, error=FALSE, message=FALSE, echo=TRUE}
knitr::opts_chunk$set(warning=FALSE, error=TRUE, message=FALSE, echo=TRUE)
fig.counter <- 1

libs  <- c("ape", "parallel", "Rcpp", "network", "sna", "GGally", "ggplot2")
invisible(lapply(libs, library, character.only=TRUE, warn.conflicts=FALSE, quietly=TRUE, verbose=FALSE))
invisible(lapply(list.files("../R", full.names=TRUE), source))
Rcpp::sourceCpp('../src/simulate.cpp')
Rcpp::sourceCpp('../src/transmission_tree.cpp')
```

## Model simulation

We can simulate an outbreak using a branching process and track who-infected-whom in order to generate the transmission tree, and thus the phylogeny of the infectious pathogen. 

In the example here, we assume an S->I->R model of disease progression in which susceptible individuals become infected and capable of infecting others, and later recover and stop being infectious. The time to recovery is exponentially distributed with rate $\gamma$. Upon recovery, an infector infects $b_{0}$ number of individuals. The number of onward infections, i.e. `offspring', caused by each infected individual is a random variable drawn from a negative binomial offspring distribution $B \sim NBin(R, k)$ with mean $R=\frac{\sum_{i=0}^{N-1}b_i}{N}$, dispersion parameter $k$, and variance $\sigma^2 = R(1+\frac{R}{k})$. The mean of the offspring distribution is the reproductive number of the infectious disease, and is related to the basic reproduction number $R_0$ via the proportion of susceptible individuals in the population: $R=R_0\frac{S}{N}$. The parameter $k$ determines the level of overdispersion in the population. At smaller values of $k$, most individuals do not cause any further infections while a few contribute to most of the transmission events. 

```{r model_params, echo=FALSE}
R0 <- 2
k <- 0.5
Tg <- 5
N <- 5000
S <- 4999
dt <- 0.1
total_dt <- 1500
min_epi_size <- 20
max_attempts <- 100
params <- c(R0=R0, k=k, Tg=Tg, N=N, S=S, I=N-S)
```

Setting $N=$ `r prettyNum(N, ",")`, $R_0=$ `r round(R0, 2)`, $k=$ `r k`, and duration of infectiousness $\frac{1}{\gamma}=$ `r Tg` days, we can simulate the outbreak using

```{r simulate_outbreak, cache=FALSE}
seed.num <- 1010113
set.seed(seed.num)
sim.outbreak <- simulate_sir(params, dt, total_dt, min_epi_size, max_attempts, TRUE)
```






The offspring distribution of the simulated epidemic follows a negative binomial distribution

`r fig.counter.offspring <- fig.counter`

```{r offspring, echo=FALSE, fig.width=4, fig.height=3, fig.cap=paste0("Figure ", fig.counter, ". The offspring distribution of the simulated epidemic assuming R0=", R0, " and k=", k, "."), cache=TRUE}
par(mar=c(5.1, 4.1, 0.25, 0.25))
offspring <- rnbinom(10000, mu=R0, size=k)
hist(offspring, xlab="Number of onward infections", main="")
legend("topright", legend=paste0("R0=", round(R0, 2), " and k=", k))
fig.counter <- fig.counter + 1
```

`r fig.counter.sim.traj <- fig.counter`

The final epidemic size was `r prettyNum(sim.outbreak$total_infected, ",")`.


## Simulated epidemic trajectories

The epidemic trajectories denoted by the incidence and prevalence curves are shown in the Figure `r fig.counter.sim.traj` below. Assuming that infectious individuals are reported at the time of recovery, the incidence curve shows the daily number of reported cases.

```{r simulated_trajectories, echo=FALSE, fig.show='hold', fig.width=7, fig.height=4, fig.cap=paste0("Figure ", fig.counter, ". The daily incidence and prevalence of the simulated epidemic. Incidence is defined as the reported cases on each day."), cache=TRUE}
par(mar=c(5.1, 4.1, 0.25, 0.25))
plot(time_series_from_line_list(sim.outbreak),
     xlab="Days since start of epidemic", ylab="Incidence per day", type="h")
plot((1:sim.outbreak$total_dt)*dt,
     sim.outbreak$prevalence,
     xlab="Days since start of epidemic", ylab="Prevalence", type="l")
fig.counter <- fig.counter+1
```

## Transmission Tree

By setting \texttt{track_transmissions} to \texttt{TRUE} we can track who infected whom in the outbreak and thus reconstruct the transmission tree. From the transmission tree, we can infer the pathogen phylogeny which describes the ancestral relationship between pathogen isolates from infected individuals.

```{r simulate_transmission_tree, echo=FALSE, cache=TRUE}
sim.transmission.tree <- as.data.frame(get_transmission_tree(sim.outbreak$infected))
sim.transmission.tree$from <- as.factor(sim.transmission.tree$from)
sim.transmission.tree$to <- as.factor(sim.transmission.tree$to)
fig.counter.sim.graph <- fig.counter
```

We can visualise the transmission network using the \texttt{get_transmission_tree} function. Below is the transmission network of the first 100 infected people.

```{r simulate_graph, echo=FALSE, fig.width=7, fig.height=7, fig.cap=paste0("Figure ", fig.counter, ". Transmission tree."), cache=TRUE}
sim.graph <- network(sim.transmission.tree[1:100, 1:2], directed=TRUE)
sim.graph %e% "length" <- sim.transmission.tree[1:100, 3]
ggnet2(sim.graph, arrow.size = 9, node.alpha=.5, label=TRUE)
fig.counter <- fig.counter + 1
```

## Phylogeny 

`r fig.counter.phylo <- fig.counter`

The phylogenetic tree is related to the transmission tree. In the case of the latter, parents are represented by internal nodes whereas in the case of phylogenies, parents are represented by an external node (tip). The \texttt{get_phylo} function produces the phylogenetic tree for a given outbreak. Figure `r fig.counter.phylo` is the phylogenetic tree of the first 100 individuals to be infected during the epidemic.

```{r simulate_tree, cache=TRUE}
tree <- get_phylo(sim.outbreak$infected)
```


```{r simulate_tree_plot, echo=FALSE, fig.width=7, fig.height=11, fig.cap=paste0("Figure ", fig.counter, ". Phylogeny of the first 100 individuals to be infected during the epidemic."), cache=TRUE}
par(mar=c(0.5, 0.5, 0.5, 0.5))
not.sampled.tips <- 101:length(tree$tip.label)
subtree <- drop.tip(tree, not.sampled.tips)
plot(subtree)
fig.counter <- fig.counter + 1
```


## Producing time-series data from simulation

Inferring parameters of dynamic disease models such as the SIR require data to be in time-series format, i.e. a quantity per time step. For epidemiologic data, this could be the number of reported cases per day. If data collected during an outbreak is in the form of a line list where each line contains information about an infected individual, this can be converted to time-series format using the function \texttt{time_series_from_line_list}. The first column should contain the ID of the infected individual and the second column the time of reporting. Here we assumed that an individual was reported upon recovery.


```{r time_series_from_line_list, cache=TRUE}
sampling.prob <- 0.01
data.dt <- 1
set.seed(seed.num)
sampled.sim.outbreak <- downsample(sim.outbreak, strategy="proportional", prob=sampling.prob)
epi_data <- time_series_from_line_list(sampled.sim.outbreak, step_size=data.dt)
head(epi_data)
```


`r fig.counter.phylo2 <- fig.counter`

And the phylogeny of the randomly sampled individuals is given in Figure `r fig.counter.phylo2`.

```{r time_series_from_tree,  echo=FALSE, fig.width=7, fig.height=11, fig.cap=paste0("Figure ", fig.counter, ". Phylogeny of ", sampled.sim.outbreak$total_sampled, " randomly sampled individuals."), cache=TRUE}
subtree <- drop.tip(reorder.phylo(tree, "postorder"), which(!(1:length(tree$tip.label) %in% sampled.sim.outbreak$sampled_individuals)))
plot(subtree, show.tip.label=FALSE)
fig.counter <- fig.counter + 1
```


```{r time_series_from_tree2}
gen_data <- time_series_from_tree(subtree, step_size=data.dt)
```


We can also obtain time-series data for both at the same time:

```{r time_series}
all_data <- get_data(epi=sampled.sim.outbreak, phy=subtree, dt=data.dt)
str(all_data)
```

## Create input files for EpiGenMCMC program

```{r create_input}
param_list <- create_params_list(
  param_names=c("R0", "k", "rateI2R", "N", "S", "reporting", "time_before_data"), # All parameter values
  init_param_values=c(R0, k, 1/Tg, N, S, sampling.prob, 10), # Initial parameter values
  params_to_estimate=c("R0", "k", "rateI2R", "reporting", "time_before_data"), # Names of parameters to be estimated
  transform=c(NA, "inverse", "inverse", NA, NA), # The algorithm will estimate the value of the transformed parameter
  prior=c("unif", "unif", "unif", "beta", "unif"), # Prior distribution
  prior_params=list(c(1.0, 100.0), c(1.0, 10000.0), c(1.0, 30.0), c(1.0, 3.0), c(0.0, 300.0)), # Parameters for the prior distribution
  proposal_params=list(c(0.5, 1.0, 100.0), c(1.0, 1.0, 10000.0), c(1.0, 1.0, 30.0), c(0.05, 0.0, 1.0), c(20.0, 0.0, 300.0))) # SD of proposal distribution, and the range of parameter values to be explored 
mcmc_options <- create_mcmc_options (particles=1000, iterations=1000, log_every=1, pfilter_every=20, 
                                     which_likelihood=0, # 0= use both epi and genetic data, 1=use only epi data, 2=use only genetic data
                                     pfilter_threshold=1.0,
                                 log_filename="log.txt", traj_filename="traj.txt")

```


